---
title: "Download_TCGAbiolinks"
author: "N. Alcala"
date: "5/6/2023"
output: 
  html_document: 
    keep_md: yes
---


# R Code to download expression data from TCGA with TCGAbiolinks

## Install TCGAbiolink
Note that due to a bug in  GDCprepare for RNAseq, it is for now necessary to install the package with these links:
```{r install, message=FALSE, eval=FALSE}
BiocManager::install("BioinformaticsFMRP/TCGAbiolinks")
BiocManager::install("BioinformaticsFMRP/TCGAbiolinksGUI.data")
```

## load libraries 
```{r libraries, message=FALSE}
library(TCGAbiolinksGUI.data)
library(TCGAbiolinks)
library(dplyr)
library(DT)
library(SummarizedExperiment)
```

## Getting the data ready
### Query platform Illumina HiSeq with a list of barcode 
We first build queries, for example for the malignant pleural mesothelioma (MESO) cohort and the breast cancer cohort (BRCA)
```{r queries, message=FALSE}
TCGA_projects = c("TCGA-ACC","TCGA-BLCA","TCGA-BRCA","TCGA-CESC","TCGA-CHOL","TCGA-COAD","TCGA-DLBC","TCGA-ESCA","TCGA-GBM","TCGA-HNSC","TCGA-KICH",
                  "TCGA-KIRC","TCGA-KIRP","TCGA-LAML","TCGA-LGG","TCGA-LIHC","TCGA-LUAD","TCGA-LUSC","TCGA-MESO","TCGA-OV","TCGA-PAAD","TCGA-PCPG",
                  "TCGA-PRAD","TCGA-READ","TCGA-SARC","TCGA-SKCM","TCGA-STAD","TCGA-TGCT","TCGA-THCA","TCGA-THYM","TCGA-UCEC","TCGA-UCS","TCGA-UVM")

queryl <- lapply(TCGA_projects, function(x) GDCquery(
  project = x, 
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
) )
```

We then download the queries
```{r download, message=FALSE, eval=FALSE}
for(x in queryl) GDCdownload(x,directory = "/data/AIML/work/alcalan/GDCdata/")
```

### Prepare expression matrix 
Matrices contain geneID in the rows and samples (barcode) in the columns, and rsem.genes.results as values
```{r prepare, message=FALSE, eval=FALSE}
for(x in queryl) GDCprepare(x,save = TRUE,directory = "/data/AIML/work/alcalan/GDCdata/")
```

### alternatively, if already prepared and saved, load from disk
```{r load, message=FALSE}
load("data/TCGA-MESOTranscriptome_ProfilingFri_Jul_29_11:39:27_2022.RData")
MESO.Rnaseq.SE = data
load("data/TCGA-BRCATranscriptome_ProfilingFri_Jul_29_12:11:29_2022.RData")
BRCA.Rnaseq.SE = data
```

### Get expression in TPM
```{r tpm, message=FALSE}
MESO.Rnaseq.expr.tpm = assay(MESO.Rnaseq.SE,"tpm_unstrand")
BRCA.Rnaseq.expr.tpm = assay(BRCA.Rnaseq.SE,"tpm_unstrand")
```

### Access clinical data
```{r clinical, message=FALSE}
colData(MESO.Rnaseq.SE)
colData(BRCA.Rnaseq.SE)
```

## Session Info 
```{r info}
sessionInfo()
```