---
title: "Download_TCGAbiolinks"
author: "N. Alcala"
date: "9/6/2022"
output: 
  html_document: 
    keep_md: yes
---


# R Code to download expression data from TCGA with TCGAbiolinks

## Install TCGAbiolink
Note that due to a bug in  GDCprepare for RNAseq, it is for now necessary to install the package with these links:
```{r install, message=FALSE, eval=FALSE}
BiocManager::install("BioinformaticsFMRP/TCGAbiolinks")
BiocManager::install("BioinformaticsFMRP/TCGAbiolinksGUI.data")
```

## load libraries 
```{r libraries, message=FALSE}
library(TCGAbiolinksGUI.data)
library(TCGAbiolinks)
library(dplyr)
library(DT)
library(SummarizedExperiment)
```

## Getting the data ready
### Query platform Illumina HiSeq with a list of barcode 
We first build queries, for example for the malignant pleural mesothelioma (MESO) cohort and the breast cancer cohort (BRCA)
```{r queries, message=FALSE}
query <- GDCquery(
  project = "TCGA-MESO", 
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)

query.BRCA <- GDCquery(
  project = "TCGA-BRCA", 
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)
```

We then download the queries
```{r download, message=FALSE, eval=FALSE}
GDCdownload(query,directory = "data/")
GDCdownload(query.BRCA,directory = "data/")
```

### Prepare expression matrix 
Matrices contain geneID in the rows and samples (barcode) in the columns, and rsem.genes.results as values
```{r prepare, message=FALSE, eval=FALSE}
MESO.Rnaseq.SE <- GDCprepare(query,save = TRUE,directory = "data/")
BRCA.Rnaseq.SE <- GDCprepare(query.BRCA,save = TRUE,directory = "data/")
```

### alternatively, if already prepared and saved, load from disk
```{r load, message=FALSE}
load("data/TCGA-MESOTranscriptome_ProfilingFri_Jul_29_11:39:27_2022.RData")
MESO.Rnaseq.SE = data
load("data/TCGA-BRCATranscriptome_ProfilingFri_Jul_29_12:11:29_2022.RData")
BRCA.Rnaseq.SE = data
```

### Get expression in TPM
```{r tpm, message=FALSE}
MESO.Rnaseq.expr.tpm = assay(MESO.Rnaseq.SE,"tpm_unstrand")
BRCA.Rnaseq.expr.tpm = assay(BRCA.Rnaseq.SE,"tpm_unstrand")
```

### Access clinical data
```{r clinical, message=FALSE}
colData(MESO.Rnaseq.SE)
colData(BRCA.Rnaseq.SE)
```

## Session Info 
```{r info}
sessionInfo()
```